---
title: "Project Setup"
author: "Kristi Lin-Rahardja"
date: "7/28/2021"
output:
  rmdformats::downcute:
    self_contained: true
    number_sections: true
    code_folding: show
    df_print: paged
  pdf_document:
    toc: yes
---

In this script, we will set up all the necessary directories and download the data from GDSC2. After this has been run once, it doesn't need to be run again.

# Setup

## Libraries

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r lib, message=FALSE}
library(here)
library(gdscIC50) #for cleaning and normalizing the raw drug response data
library(tidyverse)

conflicted::conflict_prefer("select", "dplyr")
conflicted::conflict_prefer("filter", "dplyr")
```

## Create Directories

```{r mk_directories}
#Scripts
dir.create(here("Scripts"))

#Directory where input data will be stored
dir.create(here("Data"), showWarnings = FALSE)
dir.create(here("Data", "Raw Data"), showWarnings = FALSE)
dir.create(here("Data", "Signatures"), showWarnings = FALSE)

#Directories to store results
dir.create(here("Results"), showWarnings = FALSE)

#Directories to store plots
dir.create(here("Plots"), showWarnings = FALSE)
dir.create(here("Plots", "Data Prep"), showWarnings = FALSE)
dir.create(here("Plots", "Epithelial Cancers"), showWarnings = FALSE)
```


# Download Data

To download the data from GDSC2, we will use Jessica Scarborough's code from her CisSig download_data script. The bulk of the next 3 chunks are written by her.  
The links listed in the executable chunks can be found my going to these links and right clicking/copying the specified text:  

[Expression Data](https://www.cancerrxgene.org/gdsc1000/GDSC1000_WebResources//Pathway_Activity_Scores.html)  
Copy link for "RMA normalized..."    

[Cell line Meta Data](https://www.cancerrxgene.org/downloads/bulk_download)  
Copy link for ""Cell line annotations"    

[Fitted Drug Response](https://www.cancerrxgene.org/downloads/bulk_download)  
Copy link for "GDSC2-dataset" 

[Raw Drug Response](https://www.cancerrxgene.org/downloads/bulk_download)  
Copy link for "GDSC2-raw-data"    

**If the data has been updated, the links in the following chunks need to be replaced!**
This will be obvious if an error is thrown saying the "Given file does not exist".

## Expression

In order to download the expression data, we'll create a function, `get_exp`. This function downloads the data directly from a GDSC URL, creates a zipped file in the `Data/GDSC` sub-directory of our project directory, and unzips the file. We'll read the unzipped file and store these data as a tibble in a local variable, `gdsc_expr`. These data are very large, so we need to increase the timeout limit for this download. Additionally, this function will check for the presence of the raw data file, `Cell_line_RMA_proc_basalExp.txt`, before downloading the data. If the file exists, the function will simply return the data from this file as a tibble. 

```{r load_exp, message = FALSE}
options(timeout = 600) # set the global option to stop the code if the run time exceeds 600s; increased from default of 60 s

#Create a function to download the GDSC2 expression data as a zip file from the website, then unzip it here.
get_exp <- function(cache) {
  # check if file has already been downloaded at the provided cache -- if not, proceed with the download.
  if(!file.exists(here(cache, "/Cell_line_RMA_proc_basalExp.txt"))) {
    #Download the zipped file to the specified directory
    download.file(
      "https://www.cancerrxgene.org/gdsc1000/GDSC1000_WebResources//Data/preprocessed/Cell_line_RMA_proc_basalExp.txt.zip",
      destfile = here(cache, "gdsc_expr_rma.zip"))
    #Unzip the file
    unzip(zipfile = here(cache, "gdsc_expr_rma.zip"), exdir = cache)
    #Remove the zombie zip file
    file.remove(here(cache, "gdsc_expr_rma.zip"))
  
    } else { #If the file already exists, state that we will be using the existing file.
    message("Using cached GDSC2 expression data.")
  }
  #Load the unzipped data file
  exp_df <- readr::read_tsv(paste0(cache, "/Cell_line_RMA_proc_basalExp.txt"))
  return(exp_df)
  
}

#Run the above function
raw_exprs <- get_exp(here("Data", "Raw Data"))

```

As originally loaded, the expression data set (`gdsc_expr`) has `r nrow(gdsc_expr)` rows and `r ncol(gdsc_expr)` columns. 


## Cell Line Metadata

We'll create the `get_meta` function to use `curl_download` from the `curl` package to download the cell line characterization (meta) data using file transfer protocol (FTP), storing the unaltered excel file in the `Data/GDSC` sub-directory of our project directory. We'll read this file using `read_excel` from the `readxl` package and store the data as a local variable, `cell_lines`. 

```{r load_cl}
#Create a function to download the cell line meta data/details from the website.
get_meta <- function(cache) {
   # check if file has already been downloaded at the provided cache -- if not, proceed with the download.
  if(!file.exists(paste0(cache, "/Cell_Lines_Details.xlsx"))) {
    curl::curl_download("ftp://ftp.sanger.ac.uk/pub/project/cancerrxgene/releases/current_release/Cell_Lines_Details.xlsx",
                        destfile = paste0(cache, "/Cell_Lines_Details.xlsx"))
  } else { #If the file already exists, state that we will be using the existing file.
    message("Using cached metadata.")
  }
  #Load the data file
  meta_df <- readxl::read_excel(paste0(cache, "/Cell_Lines_Details.xlsx"))
  return(meta_df)
}

#Run the above function
cell_lines <- get_meta(here("Data", "Raw Data"))
```

As originally loaded, the cell line meta data set (`data_raw_meta`) has `r nrow(data_raw_meta)` rows and `r ncol(data_raw_meta)` columns. 





## Fitted Drug Response

For the drug response data, we'll create the `get_dr` function, using `curl_download` from the `curl` package to download the drug response data using file transfer protocol (FTP), storing the unaltered excel file in the `Data/GDSC` sub-directory of our project directory. We'll read this file using `read_excel` from the `readxl` package and store the data as a local variable, `gdsc_dr`. 

```{r load_fit_dr}
#Create a function to download the GDSC2 drug response datafrom the website.
get_dr <- function(cache) {
   # check if file has already been downloaded at the provided cache -- if not, proceed with the download.
  if(!file.exists(paste0(cache, "/GDSC2_fitted_dose_response.xlsx"))) {
    # Download file, store in specified directory
    curl::curl_download(
      "https://cog.sanger.ac.uk/cancerrxgene/GDSC_release8.5/GDSC2_fitted_dose_response_27Oct23.xlsx",
      destfile = paste0(cache, "/GDSC2_fitted_dose_response.xlsx"))
    
  } else { #If the file already exists, state that we will be using the existing file.
    message("Using cached GDSC2 drug response data.")
  }
  
  #Load the data file
  dr_df <- readxl::read_excel(paste0(cache, "/GDSC2_fitted_dose_response.xlsx"))
  return(dr_df)
  
}

#Run the above function
fitted_dr <- get_dr(here("Data", "Raw Data"))
```

As originally loaded, the drug response data set (`gdsc_dr`) has `r nrow(gdsc_dr)` rows and `r ncol(gdsc_dr)` columns.


## Raw Drug Response

For the raw drug response data, we'll create the `get_raw_dr` function, using `curl_download` from the `curl` package to download the drug response data using file transfer protocol (FTP), storing the unaltered excel file in the `Data/GDSC` sub-directory of our project directory. We'll read this file using `read_csv` and store the data as a local variable, `raw_dose_resp`. 

```{r load_dr}
#Create a function to download the GDSC2 drug response datafrom the website.
get_raw_dr <- function(cache) {
   # check if file has already been downloaded at the provided cache -- if not, proceed with the download.
  if(!file.exists(paste0(cache, "/gdsc2_raw_drug_response.csv"))) {
    # Download file, store in specified directory
    curl::curl_download(
      "https://cog.sanger.ac.uk/cancerrxgene/GDSC_release8.5/GDSC2_public_raw_data_27Oct23.csv.zip",
      destfile = paste0(cache, "/gdsc2_raw_drug_response.csv"))
    
  } else { #If the file already exists, state that we will be using the existing file.
    message("Using cached GDSC2 raw drug response data.")
  }
  
  #Load the data file
  dr_df <- read_csv(paste0(cache, "/gdsc2_raw_drug_response.csv"))
  return(dr_df)
  
}

#Run the above function
raw_dose_resp <- get_raw_dr(here("Data", "Raw Data"))

```



# Clean Data

## Load Data (if not already loaded)

```{r}
cell_lines = readxl::read_xlsx(here("Data", "Raw Data", "Cell_Lines_Details.xlsx"))
raw_exprs = readr::read_tsv(here("Data", "Raw Data", "Cell_line_RMA_proc_basalExp.txt"))
fitted_dr = readxl::read_excel(here("Data", "Raw Data", "GDSC2_fitted_dose_response.xlsx"))
raw_dose_resp = read_csv(here("Data", "Raw Data", "gdsc2_raw_drug_response.csv"))
```


## Cell Line Metadata

```{r clean_cl}
#Make this column name easier to reference
colnames(cell_lines)[c(9,10)] = c("Tissue_Descriptor", "Subtype_Label")

#Rename COAD/READ and unclassified so the labels match between dfs
cell_lines$Subtype_Label = cell_lines$Subtype_Label %>% replace_na('UNCLASSIFIED') #for cell lines w/o classifications, we'll label as unclassified
cell_lines[cell_lines == "UNABLE TO CLASSIFY"] = 'UNCLASSIFIED'
cell_lines[cell_lines == "COAD/READ"] = 'COREAD'

#many rows have tissue_desc and a subtype label exists, but is NA in that column--how to replace the NAs? cl_key?
label_key = cell_lines %>% select(Tissue_Descriptor, Subtype_Label)
label_key = label_key[!duplicated(label_key),]

#List out all the tissues that have duplicate rows, one NA and one properly labeled
tissue_list = unique(label_key$Tissue_Descriptor)
for (i in 1:length(tissue_list)){
  if(length(which(label_key$Tissue_Descriptor == tissue_list[i])) > 1){
    print(tissue_list[i])
  }
}

cell_lines$test = cell_lines$Subtype_Label
cell_lines$test = str_replace(cell_lines$Tissue_Descriptor, "head and neck", "HNSC")
cell_lines$test = str_replace(cell_lines$Tissue_Descriptor, "haematopoietic_neoplasm other", "MM")
cell_lines$test = str_replace(cell_lines$Tissue_Descriptor, "lymphoblastic_leukemia", "ALL")
cell_lines$test = str_replace(cell_lines$Tissue_Descriptor, "T_cell_leukemia", "ALL")
cell_lines$test = str_replace(cell_lines$Tissue_Descriptor, "breast", "BRCA")
cell_lines$test = str_replace(cell_lines$Tissue_Descriptor, "stomach", "STAD")
cell_lines$test = str_replace(cell_lines$Tissue_Descriptor, "kidney", "KIRC")
cell_lines$test = str_replace(cell_lines$Tissue_Descriptor, "lung_NSCLC_adenocarcinoma", "LUAD")
cell_lines$test = str_replace(cell_lines$Tissue_Descriptor, "pancreas", "PAAD")
cell_lines$test = str_replace(cell_lines$Tissue_Descriptor, "cervix", "CESC")
cell_lines$test = str_replace(cell_lines$Tissue_Descriptor, "endometrium", "UCEC")
cell_lines$test = str_replace(cell_lines$Tissue_Descriptor, "ovary", "OV")
cell_lines$test = str_replace(cell_lines$Tissue_Descriptor, "prostate", "PRAD")
```

```{r save_cl}
#Save the cleaned, full metadata
saveRDS(cell_lines, here("Data", "cleaned_cell_line_meta.rds"))
```



```{r subtype_key}
#create a key to match cell lines to their subtypes
subtype_key = cell_lines %>% select(`COSMIC identifier`, Subtype_Label)
subtypes = unique(subtype_key$Subtype_Label)
colnames(subtype_key) = c("cell_line", "subtype")
class(subtype_key$cell_line) = "character"
```

```{r save_subtype_key}
saveRDS(subtype_key, here("Data", "subtype_key.rds"))
```


## Expression

Tidy up the data a bit by cleaning the column names and removing duplicate genes.

```{r clean_exp}
#Change the column names from DATA.COSMIC ID to just COSMIC IDs (COSMIC = cell lines)
cols = colnames(raw_exprs) #store the column names as a vector
cols = as.data.frame(cols[-c(1:2)]) #Change cols vector to a data frame and remove the first two descriptor columns
colnames(cols) = "x" #the name of the column in cols is weird to reference, just change it to "x" for ease of reference
cols = cols %>%
  separate(x, c("rm", "COSMIC_ID"), "DATA.") #split the values-- create 2 new columns: one with "DATA." and the other with the cosmic ID
cols = as.vector(cols$COSMIC_ID) #make the cols variable a vector of just the cosmic IDs
colnames(raw_exprs)[3:ncol(raw_exprs)] = cols #Set the column names of raw_exprs (minus the 2 descriptor columns) to be just the IDs
rm(cols) #Remove this variable now that we're done with it for cleanliness

#check that there arent duplicated genes- output should be NA and nothing else. If there is more, use the commented line below
#many of the duplicated genes are just "NA"s; rename them so R doesnt get confused
raw_exprs$GENE_SYMBOLS = raw_exprs$GENE_SYMBOLS %>% replace_na('UNCLASSIFIED')
dup_genes_gdsc = raw_exprs$GENE_SYMBOLS[c(which(duplicated(raw_exprs$GENE_SYMBOLS)==TRUE))]
```

Normalize the expression.

```{r norm_exp}
#Normalize expression using scale(), by genes (columns)
exprs_norm = raw_exprs #Copy the original data
exprs_norm[,3:ncol(exprs_norm)] = scale(exprs_norm[,3:ncol(exprs_norm)],
                                        center=TRUE,
                                        scale=TRUE) #center = subtract by the mean; scale = divide by stdev


#Clean + reformat a bit
exprs_norm = exprs_norm %>%
  dplyr::select(-GENE_title) %>% #remove the extra descriptor column
  filter(!(GENE_SYMBOLS %in% dup_genes_gdsc)) %>% # remove duplicated genes
  mutate(GENE_SYMBOLS = str_replace_all(GENE_SYMBOLS, "-", "_")) %>% #replace any - with _
  gather(COSMIC_ID, value, -GENE_SYMBOLS) %>% #convert to long format (excluding the gene symbols col)
  spread(GENE_SYMBOLS, value) #convert back to wide format, with genes as colnames
```

Save the cleaned and normalized expression data.

```{r save_exp}
#Save
saveRDS(exprs_norm, here("Data", "normalized_exprs.rds"))
```



Try normalizing using a set of housekeeping genes instead of the 2 chunks above (z-scores).

```{r}
#check exprs distribution of a few housekeeping genes of interest
plot_exprs = pivot_longer(raw_exprs,
                          names_to = "COSMIC_ID",
                          values_to = "raw_exprs",
                          cols=-c("GENE_SYMBOLS", "GENE_title"))

library(patchwork)
(ggplot(plot_exprs %>% filter(GENE_SYMBOLS=="GAPDH"), aes(y=raw_exprs)) +
  geom_boxplot(width=0.1) + 
  scale_y_continuous(limits=c(min(plot_exprs$raw_exprs[which(plot_exprs$GENE_SYMBOLS=="GAPDH")]), 13)) +
  labs(title="GAPDH Raw Expression")) 
# (ggplot(plot_exprs %>% filter(GENE_SYMBOLS=="ACTB"), aes(y=raw_exprs)) +
#   geom_boxplot() + 
#    scale_y_continuous(limits=c(min(plot_exprs$raw_exprs[which(plot_exprs$GENE_SYMBOLS=="ACTB")]), 13)) +
#   labs(title="ACTB Raw Expression"))+
# (ggplot(plot_exprs %>% filter(GENE_SYMBOLS=="RPL13A"), aes(y=raw_exprs)) +
#   geom_boxplot() + 
#    scale_y_continuous(limits=c(min(plot_exprs$raw_exprs[which(plot_exprs$GENE_SYMBOLS=="RPL13A")]), 13)) +
#   labs(title="RPL13A Raw Expression")) +
# (ggplot(plot_exprs %>% filter(GENE_SYMBOLS=="TUBB"), aes(y=raw_exprs)) +
#   geom_boxplot() + 
#    scale_y_continuous(limits=c(min(plot_exprs$raw_exprs[which(plot_exprs$GENE_SYMBOLS=="TUBB")]), 13)) +
#   labs(title="TUBB Raw Expression"))+
# (ggplot(plot_exprs %>% filter(GENE_SYMBOLS=="APOL3"), aes(y=raw_exprs)) +
#   geom_boxplot() + 
#    scale_y_continuous(limits=c(min(plot_exprs$raw_exprs[which(plot_exprs$GENE_SYMBOLS=="APOL3")]), 13)) +
#   labs(title="APOL3 Raw Expression"))

(ggplot(plot_exprs, aes(y=raw_exprs, x="All Genes")) +
  geom_violin(fill="#cad2c5", col = "#2f3e46") +
  geom_boxplot(width=0.05, fill="#52796f", col="#2f3e46") +
   
  geom_violin(data = plot_exprs %>% filter(GENE_SYMBOLS=="GAPDH"), aes(y=raw_exprs, x="GAPDH"),
              fill="#cad2c5", col = "#2f3e46") +
  geom_boxplot(data = plot_exprs %>% filter(GENE_SYMBOLS=="GAPDH"), aes(y=raw_exprs, x="GAPDH"),
    width=0.05, fill="#52796f", col="#2f3e46") +
   
  scale_y_continuous(limits=c(min(plot_exprs$raw_exprs), 14)) +
  labs(title="Raw Expression of All Genes vs GAPDH",
       x="Gene(s)",
       y="Raw Expression Value"))

ggsave(here("Plots" ,"Data Prep", "exprs_dist_gapdh.png"),
       width=8, height=6)

range(plot_exprs %>% filter(GENE_SYMBOLS=="GAPDH")%>%select(raw_exprs))
range(plot_exprs %>% filter(GENE_SYMBOLS=="ACTB")%>%select(raw_exprs))
range(plot_exprs %>% filter(GENE_SYMBOLS=="RPL13A")%>%select(raw_exprs))
range(plot_exprs %>% filter(GENE_SYMBOLS=="TUBB")%>%select(raw_exprs))
range(plot_exprs %>% filter(GENE_SYMBOLS=="APOL3")%>%select(raw_exprs))
range(plot_exprs$raw_exprs)


#gapdh has the smallest range/is most consistently high, so let's try using just that one
```

```{r}
ctrl = raw_exprs[which(raw_exprs$GENE_SYMBOLS=="GAPDH"),] #select just the gapdh exprs values

norm = raw_exprs #new df to replace w normalized vals
unique(colnames(norm) == colnames(ctrl)) #should just be TRUE

for(i in 3:ncol(raw_exprs)){
  norm[,i] = norm[,i]/as.numeric(ctrl[,i]) #normalize all genes within samples
}

unique(as.vector(norm %>% filter(GENE_SYMBOLS=="GAPDH"))) #should only have 2 genes names and the number 1

plot_exprs_norm = pivot_longer(norm, #pivot to check
                          names_to = "COSMIC_ID",
                          values_to = "norm_exprs",
                          cols=-c("GENE_SYMBOLS", "GENE_title"))
range(plot_exprs_norm %>% filter(GENE_SYMBOLS=="GAPDH")%>%select(norm_exprs)) #compare ranges again
range(plot_exprs_norm %>%select(norm_exprs))

exprs_norm2 = pivot_wider(plot_exprs_norm %>% filter(GENE_SYMBOLS != "UNCLASSIFIED"), #format the same as the other norm exprs df
                          values_from = "norm_exprs",
                          names_from = "GENE_SYMBOLS",
                          id_cols="COSMIC_ID")

exprs_norm2 = exprs_norm2 %>% #order the columns alphanetically like the other norm exprs df
  select(COSMIC_ID, sort(colnames(exprs_norm2[,-1])))
```

```{r}
#Save
saveRDS(exprs_norm2, here("Data", "normalized_exprs_to_gapdh.rds"))
```


## Fitted Drug Response

Make a key to decipher the drug IDs using the fitted drug response dataset.

```{r drug_id_key}
drug_id_key = fitted_dr %>% select(DRUG_ID, DRUG_NAME)

saveRDS(drug_id_key, here("Data", "drug_id_key.rds"))
```



## Raw Drug Response

The following 2 chunks are based on the `gdscIC50` vignette linked [here](https://rdrr.io/github/yemelianovskyi/gdscIC50_advanced/f/vignettes/gdscIC50.Rmd).

```{r clean_dr}
#Remove any drug treatments that were failed. Then, remove any sets with no indicated drug.
cleaned_dose_resp = removeFailedDrugs(raw_dose_resp) %>% 
  removeMissingDrugs()
```

```{r norm_dr}
#Convert raw fluorescence intensities to cell viability/survival (between 0 and 1); normalize per plate
norm_dose_resp = normalizeData(cleaned_dose_resp,
                              trim = T, #ensure no values are != 1 or != 0
                              neg_control = "NC-1",
                              pos_control = "B")
```

```{r save_dr}
saveRDS(norm_dose_resp, here("Data", "prepped_drug_resp.rds"))
```


# Session Info

```{r}
sessionInfo()
```



